---
title : "Applications of Probability Theory"
author: "Pinak Shome"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(HistData)
```


# Load Data

```{r}
data("PolioTrials")
dat<-PolioTrials
```


## Context

Polio Trails last part in this repository addresses the question of whether the `NotInoculated` and `Placebo` groups in the `Randomized Control` experiment had statistically significantly different rates of paralytic polio. 

Recall that the `NotInoculated` and `Placebo` groups differ in that the children in the `Placebo` group had been enrolled in the vaccine trial while the parents of the children in the `NotInoculated` group did not enroll their children. 

The approach, using the `rbinom` function, implemented the idea that populations in the `NotInoculated` and `Placebo` groups in the `RandomizedControl` experiment were the same in regards to paralytic polio cases by using the `rbinom` function to assign paralytic polio cases in the combined `NotInoculated` and `Placebo` groups of the `RandomizedControl` experiment to the `Placebo` group with probability equal to the ratio of the size of the `Placebo` group to the size of pooled `Placebo` group and `NotInoculated` group.

Note that the function `rbinom(x,size,prob)` simulates drawing $x$ random samples from `Binom(size,prob)`. 

The computations for that analysis are reproduced here:

```{r}
n<-10000 # number of simulations

# Calculate the number of paralytic polio cases in the pooled "Placebo" and "NotInoculated" group.
ct<-sum(dat$Paralytic[2:3])

# Calculate the proportion "prop" of the the pooled "Placebo" and "NotInoculated" group that are in the "Placebo" group.
prop<-dat$Population[2]/sum(dat$Population[2:3])

# Generate 10,000 counts of paralytic polio cases in the "Placebo" group under the model that each paralytic polio case in the pooled pooled "Placebo" and "NotInoculated" group has probability "prop" of being assigned to the "Placebo" group.
set.seed(45678765)
sim2<-rbinom(n,ct,prop) 
qplot(sim2,bins=20)
# proportion of the simulated counts of paralytic polio in the "Placebo" that are less than or equal to the observed count:
mean(sim2<=dat$Paralytic[2])
# proportion of the simulated counts of paralytic polio in the "Placebo" that are greater than or equal to the observed count:
mean(sim2>=dat$Paralytic[2])
```

##part 1

Using the same null model described above, we will calculate the probability that the count of paralytic polio cases in the `Placebo` group under the null model is less than or equal to  `dat$Paralytic[2]` directly rather than by simulating it. Recall that the function `pbinom(x,size,prob)` returns the probability of the event that the number of successes is in the set $\{0,1,...x\}$.


```{r}
A <- pbinom(dat$Paralytic[2], ct, prop)
A
## [1] 0.9998718

```

## part 2

Using the same null model described above, we will calculate the probability that the count of paralytic polio cases in the `Placebo` group under the null model is greater than or equal to `dat$Paralytic[2]` directly rather than by simulating it.

```{r}
B <- pbinom(dat$Paralytic[2] - 1, ct, prop)
1-B
## [1] 0.0002119741
```

##part 3

Observations:

This probability can be interpreted as a p-value in the context of hypothesis testing. A low p-value (commonly below 0.05) indicates that the observed result is unlikely under the null hypothesis. In other words, it would suggest that the observed number of cases in the Placebo group is statistically significantly higher than what the null model would predict. This could be evidence against the null hypothesis, leading to its rejection in favor of an alternative hypothesis that there is a difference between the groups.